---
title: "Topics on Welfare Policy"
author: "Danton Noriega"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE, fig.height = 4.6, fig.width = 7.6, dev = "svg")
options(htmltools.dir.version = FALSE)
rfiles <- list.files("R")
invisible(sapply(file.path("R", rfiles), source))
```

# At a Glance

### *Chapter 1* <br/>An Analysis of the Double Up Food Bucks Program
### *Chapter 2* <br/>The Durham Connects Pilot & Requests for Social Services
### *Chapter 3* <br/>SNAP EBT Transfer Dates & Off-Cycle Demand for Formal Assistance


<!-- CHAPTER 1 -->


---
class: center, middle, inverse

# Chapter 1

## Double Up Food Bucks 

---

### Research Question

> How effective is the DUFB financial incentive at increasing the fresh fruits and vegetables purchased by SNAP shoppers within a grocery store environment?

--

### Hypotheses

- A small increase (2 - 5%) in SNAP EBT expenditures on fruits and vegetables (FVs) during the months the DUFB incentive is active in the experimental stores (Aug - Dec).
- No significant differences in FV expenditures during the months the incentive is not in effect (Jan - July).
- The state's SNAP benefit schedule will affect when differences are detected. Difference, if found, will be detected during the 1st, 2nd, and 3rd weeks, fading during the 4th week.

---
# Concept in a Plot

```{r echo=FALSE}
source("R/DD.R")
DD1()
```

.footnote[
> FAKE data.
]


---

### Motivation

- High consumption of unhealthy foods, low consumption of fruits and vegetables
- Overtime, poor diet leads to adverse health outcomes
- Chronic diet-related diseases are expensive
- US government could save lives and money by improving diets

--

### Supplemental Nutrition Assistance Program (SNAP)

- Anti-hunger program, not a nutrition program
- Currently, few restrictions on what can be purchased with SNAP benefits
- Substantial overlap between families participating in SNAP and Medicaid/Medicare
- These programs are expensive
    - SNAP — $74 billion FY2015
    - Medicare — $646.2 billion FY2015
    - Medicaid - $545.1 billion FY2015

*Possible to save money by improving the diet of SNAP beneficiaries.*

---

### Double Up Food Bucks (DUFB)

- Financial incentive program targetting SNAP beneficiaries
- Doubles purchasing power for FVs up to $20/day
- Dollars (points) are earned buying local fresh produce
- Can then be redeemed on *any* eligible produce

--

### Data

- 3 years of sales data from 32 stores (17 treated, 15 control)
- Sales data includes UPC, price, tender, time, etc.
- **Status**: Testing transfer options.

--

### Known and Expected Data Challenges

1. Purchases not linked to customers
2. Outcome Variable with non-trivial amount of zeros (corner solutions)
3. Consumers maximize across multiple product types

---

### Methods

All cross-sectional.

1. Difference-in-Difference-in-Differences
2. Hurdle Models
3. Multiple Discrete-Continuous Extreme Value

--

### Unobserved Individual Effects?

*Working on it.*

- Assume shoppers fall into types 
  - but also assume shopper type is dynamic
- Determine shopper types via clustering using 
  - shopping basket composition
  - time/day of purchase





<!-- CHAPTER 2 -->

---
class: center, middle, inverse

# Chapter 2

## Durham Connects & Request for Social Services

---

### Research Question: 

> For families that participated in the Durham Connects pilot program, did the intervention (information and nurse contact) impact the future probability of applying for social services?

--

### Hypothesis

- Expect to observed a larger proportion of families that participated in Durham Connects requesting social services. 
- Hypothesis is that the information and nurse contact provided to DC families decreased the complexity of applying and navigating the social safety net. 
- DC families are observed more frequently not because they are, on average, in greater need of assistance, but because they have been informed, encouraged, and enabled to seek out available resources.

---
# Concept in a Plot

```{r echo=FALSE}
set.seed(1985)
y1 <- purrr::map_dbl(1 + rnorm(300, 1,2), round) %>%
  data_frame(Apps = ., DC = "No")
y2 <- purrr::map_dbl(1 + rnorm(300, 2,2), round) %>%
  data_frame(Apps = ., DC = "Yes")
d <- bind_rows(y1, y2) %>%
    dplyr::mutate(Apps = replace(Apps, Apps < 0, NA))

clrs <- scales::hue_pal()(2)
ns <- d %>% 
  na.omit() %>%
  group_by(DC) %>%
  summarise(n = n())

n1 <- ns$n[1]
n2 <- ns$n[2]


ggplot2::ggplot(data = d, aes(x=Apps, fill = DC, alpha = DC)) +
  scale_alpha_manual(values = c(.8, .5)) +
  geom_histogram(position = 'dodge', binwidth = .5) + 
  # stat_function(fun = function(x, mean, sd, n, lower, bw){ 
  #     msm::dtnorm(x = x, mean = mean, sd = sd, lower = lower) * n * bw
  #   }, 
  #   args = list(mean = 2, sd = 2, lower = 0, n = n1 + n2, bw = .5),
  #   color = clrs[1], size = 3, alpha = .8) +
  # stat_function(fun = function(x, mean, sd, n, lower, bw){ 
  #     msm::dtnorm(x = x, mean = mean, sd = sd, lower = lower) * n * bw
  #   }, 
  #   args = list(mean = 3, sd = 2, lower = 0, n = n1 + n2, bw = .5),
  #   color = clrs[2], size = 3, alpha = .5) +
  theme_clean() +
  labs(x = expression(paste("Observed Applications in ",
            group("(",list(0,T),"]"))),
    y = "Count", title = "Observed Social Service Applications (T years since birth)")
```


---

### Take-Up Rate

> Percentage of eligible individuals receiving benefits.

--

### Take-Up Rate can be reduced by

- Stigma
- Onerous application process
- Confusion, misunderstanding, misinformation

### Frameworks

- **Neoclassical**: rational, complete information
  - Gap explained by *stigma* and other costs.
- **Bounded Rationality**: incomplete information, hueristics, status quo bias


---

# Simple Model

.pull-left[
$b$ = Benefit
$d$ = Decision to apply
]

.pull-right[
$c$ = Cost of applying
$\tilde c$ = Cost of applying *after DC* 
]










<!-- CHAPTER 2 -->

---
class: center, middle, inverse

# Chapter 3

## SNAP and Off-Cycle Demand for Assistance

---

### Research Question

> Do request for additional assistance increase during the off-cycles of a household SNAP benefit schedules?

### Hypothesis

- Expect, in aggregate, an increase towards the last 2 weeks of the month following SNAP benefit transfers. 
- That is, I expect to reject the null hypothesis that, in aggregate, request for formal assistance are randomly distributed between transfer dates.

---

# Concept in a Plot

```{r echo=FALSE}
set.seed(1985)
dr <- purrr::map(1:30, ~data_frame(t=.x, y=rnorm(12,4,3), h="H0")) %>%
  dplyr::bind_rows()
dnr <- purrr::map(1:30, ~data_frame(t=.x, y=.008*.x^2 + rnorm(12,3,3), h="H1")) %>%
  dplyr::bind_rows()
d <- bind_rows(dr, dnr) %>%
    dplyr::mutate(y = replace(y, y < 0, 0))
ggplot2::ggplot(data = d, aes(x=t,y=y,color = h)) + 
  geom_point(size=1) + 
  scale_size_manual(values=c(2,3)) +
  geom_smooth() + 
  theme_clean() +
  labs(y = "Requests", x = "Days since Transfer", title = "Pooled Request by Day since Transfer")
```

> FAKE data.







