---
title: "Topics on Welfare Policy"
author: "Danton Noriega"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE, fig.height = 4.6, fig.width = 7.6, dev = "svg")
options(htmltools.dir.version = FALSE)
rfiles <- list.files("R")
invisible(sapply(file.path("R", rfiles), source))
```

# At a Glance

### Chapter 1 <br/>*An Analysis of the Double Up Food Bucks Program*
### Chapter 2 <br/>*The Durham Connects Pilot & Requests for Social Services*
### Chapter 3 <br/>*SNAP EBT Transfer Dates & Off-Cycle Demand for Formal Assistance*


<!-- CHAPTER 1 -->


---
class: center, middle, inverse

# Chapter 1

## Double Up Food Bucks 

---

### Research Question

> How effective is the DUFB financial incentive at increasing the fresh fruits and vegetables purchased by SNAP shoppers within a grocery store environment?

--

### Hypotheses

- A small increase (2 - 5%) in SNAP EBT expenditures on fruits and vegetables (FVs) during the months the DUFB incentive is active in the experimental stores (Aug - Dec).
- No significant differences in FV expenditures during the months the incentive is not in effect (Jan - July).
- The state's SNAP benefit schedule will affect when differences are detected. Difference, if found, will be detected during the 1st, 2nd, and 3rd weeks, fading during the 4th week.

---
# Concept in a Plot

```{r echo=FALSE}
source("R/DD.R")
DD1()
```

.footnote[
> FAKE data.
]


---

### Motivation

- High consumption of unhealthy foods, low consumption of fruits and vegetables
- Overtime, poor diet leads to adverse health outcomes
- Chronic diet-related diseases are expensive
- US government could save lives and money by improving diets

--

### Supplemental Nutrition Assistance Program (SNAP)

- Anti-hunger program, not a nutrition program
- Currently, few restrictions on what can be purchased with SNAP benefits
- Substantial overlap between families participating in SNAP and Medicaid/Medicare
- These programs are expensive
    - SNAP — $74 billion FY2015
    - Medicare — $646.2 billion FY2015
    - Medicaid - $545.1 billion FY2015

*Possible to save money by improving the diet of SNAP beneficiaries.*

---

### Double Up Food Bucks (DUFB)

- Financial incentive program targetting SNAP beneficiaries
- Doubles purchasing power for FVs up to $20/day
- Dollars (points) are earned buying local fresh produce
- Can then be redeemed on *any* eligible produce

--

### Data

- 3 years of sales data from 32 stores (17 treated, 15 control)
- Sales data includes UPC, price, tender, time, etc.
- **Status**: Testing transfer options.

--

### Known and Expected Data Challenges

1. Purchases not linked to customers
2. Outcome Variable with non-trivial amount of zeros (corner solutions)
3. Consumers maximize across multiple product types

---

### Methods

All cross-sectional.

1. Difference-in-Difference-in-Differences
2. Hurdle Models
3. Multiple Discrete-Continuous Extreme Value

--

### Unobserved Individual Effects?

*Working on it.*

- Assume shoppers fall into types 
  - but also assume shopper type is dynamic
- Determine shopper types via clustering using 
  - shopping basket composition
  - time/day of purchase





<!-- CHAPTER 2 -->

---
class: center, middle, inverse

# Chapter 2

## Durham Connects & Request for Social Services

---

### Research Question: 

> For families that participated in the Durham Connects pilot program, did the intervention (information and nurse contact) impact the future probability of applying for social services?

--

### Hypothesis

- Expect to observed a larger proportion of families that participated in Durham Connects requesting social services. 
- Hypothesis is that the information and nurse contact provided to DC families decreased the complexity of applying and navigating the social safety net. 
- DC families are observed more frequently not because they are, on average, in greater need of assistance, but because they have been informed, encouraged, and enabled to seek out available resources.

---
# Concept in a Plot

```{r echo=FALSE}
set.seed(1985)
y1 <- purrr::map_dbl(1 + rnorm(300, 1,2), round) %>%
  data_frame(Apps = ., DC = "No")
y2 <- purrr::map_dbl(1 + rnorm(300, 2,2), round) %>%
  data_frame(Apps = ., DC = "Yes")
d <- bind_rows(y1, y2) %>%
    dplyr::mutate(Apps = replace(Apps, Apps < 0, NA))

clrs <- scales::hue_pal()(2)
ns <- d %>% 
  na.omit() %>%
  group_by(DC) %>%
  summarise(n = n())

n1 <- ns$n[1]
n2 <- ns$n[2]


ggplot2::ggplot(data = d, aes(x=Apps, fill = DC, alpha = DC)) +
  scale_alpha_manual(values = c(.8, .5)) +
  geom_histogram(position = 'dodge', binwidth = .5) + 
  # stat_function(fun = function(x, mean, sd, n, lower, bw){ 
  #     msm::dtnorm(x = x, mean = mean, sd = sd, lower = lower) * n * bw
  #   }, 
  #   args = list(mean = 2, sd = 2, lower = 0, n = n1 + n2, bw = .5),
  #   color = clrs[1], size = 3, alpha = .8) +
  # stat_function(fun = function(x, mean, sd, n, lower, bw){ 
  #     msm::dtnorm(x = x, mean = mean, sd = sd, lower = lower) * n * bw
  #   }, 
  #   args = list(mean = 3, sd = 2, lower = 0, n = n1 + n2, bw = .5),
  #   color = clrs[2], size = 3, alpha = .5) +
  theme_clean() +
  labs(x = expression(paste("Observed Applications in ",
            group("(",list(0,T),"]"))),
    y = "Count", title = "Observed Social Service Applications (T years since birth)")
```

---

# Durham Connects (DC) Program

- Piloted from July 1, 2009 to Dec 31, 2010
  - 4777 resident births randomly assigned to DC by DOB (even days)
- Universal postnatal nurse home-visiting program
- Aim to reduce emergency room visits and reduce child maltreatment
- Nurses assess and rate 12 factors along 4 domains
  - Domains: *Health Care, Caring for Infant, Home Safety, Parent Wellbeing*
  - Rated from 1 to 4
      - $1$ - no risk, basic guidance provided
      - $2$ - needs/concerns addressed via education & guidance
      - $3$ - issues addressed by linkage with community resources
      - $4$ - urgent need, immediate invervention required
- Top referral was to Durham Social Service (Erica Ward)
  

---

### Take-Up Rate

> Percentage of eligible individuals receiving benefits.

--

### Take-Up Rate can be reduced by

- Stigma
- Onerous application process
- Confusion, misunderstanding, misinformation

--

### Economic Frameworks

- **Neoclassical**: rational, complete information
  - Gap explained by *stigma* and other costs (search, transaction).
- **Bounded Rationality**: incomplete information, hueristics, status quo bias

---

## Data

### Center for Child and Family Health (CCFH)

- Durham Connects Pilot Data
  - **Short-term birth certificates**
  - More detailed DC data exists
      - IRB requirements complicated
      - Don't think value added justifies efforts for access
--

### Durham Social Services (DSS)

- Administrative Records
  - Applications for Social Services (2012 - 2016)
      - Verifying if data from 2009 - 2012 exists
  - Also have SNAP, Medicaid, TANF
  - Do not have WIC
      - Feasible

---

# Simple Model

Assume all individuals are the same *except for assignment to DC*.
<hr>

$$\begin{gathered}
  r  = \left \{ 
    \begin{array}{c} 
      1 \text{ if } NB > 0 \\ 
      0 \text{ if } NB \le 0 
    \end{array} \right. \\
   \\
  NB = B - c \\
  \\
P(d=1|\text{DC}) > P(d=1| \neg \text{DC})
\end{gathered}$$

<hr>

.pull-left[
$B~~~$ = Benefit <br>
$NB$ = Net Benefit <br>
$r~~~$ = Decision to apply (participate)
]

.pull-right[
$c$ = Cost of applying <br>
$\tilde c$ = Cost of applying *given DC* <br>
$\quad~$ where $c > \tilde c$
]

---

# Regression Framework

Of interest:
1. Probability of being observed applying *at all* (participation)
2. How frequently observed applying (demand/consumption)

--

<hr>
Let $r$ represent the decision to participate and $D$ assignment to DC.
Observed applications (count), $y$, since birth of child for some fixed time $T$.
<hr>

--

$$\begin{gather}
w^* = \mathbf{z'} \alpha + \delta D + u, \quad
r = \left \{ 
  \begin{array} 
    ~1 \text{ if } w^* > 0 \\ 
    ~0 \text{ otherwise}
  \end{array} \right .\\
y^* = f(\mathbf{x}, D, v), \quad
y = r \cdot max(0, y^*) \quad
\end{gather}$$
$$\begin{aligned}
E[y^*|\mathbf{x},D] &= \mathbf{x' \beta} + \gamma D \quad \quad \quad \text{(OLS)} \\
E[y^*|\mathbf{x},D] &= exp(\mathbf{x' \beta} + \gamma D) \quad \text{(Poisson)}
\end{aligned}$$
$$\star ~ E[u|D] = E[v|D] = 0 ~ \star$$




<!-- CHAPTER 2 -->

---
class: center, middle, inverse

# Chapter 3

## SNAP and Off-Cycle Demand for Assistance

---

### Research Question

> Do request for additional assistance increase during the off-cycles of a household SNAP benefit schedules?

### Hypothesis

- Expect, in aggregate, an increase towards the last 2 weeks of the month following SNAP benefit transfers. 
- That is, I expect to reject the null hypothesis that, in aggregate, request for formal assistance are randomly distributed between transfer dates.

---

# Concept in a Plot

```{r echo=FALSE}
set.seed(1985)
dr <- purrr::map(1:30, ~data_frame(t=.x, y=rnorm(12,4,3), h="H0")) %>%
  dplyr::bind_rows()
dnr <- purrr::map(1:30, ~data_frame(t=.x, y=.008*.x^2 + rnorm(12,3,3), h="H1")) %>%
  dplyr::bind_rows()
d <- bind_rows(dr, dnr) %>%
  dplyr::mutate(y = replace(y, y < 0, 0)) %>%
  dplyr::mutate(y = round(y))  
ggplot2::ggplot(data = d, aes(x=t,y=y,color = h)) + 
  geom_jitter(size=1, width = .05) + 
  scale_size_manual(values=c(2,3)) +
  geom_smooth() + 
  theme_clean() +
  labs(y = "Requests", x = "Days since Transfer", title = "Pooled Request by Day since Transfer (12 Months)")
```

> FAKE data.

---

### SNAP Overview

- Cash-like benefit
- Popular, expansive
- Spent quickly
- Not enough

--

### SNAP Benefit Schedule (North Carolina)

- On-Cycle
  - All odd days between 3 and 21, mapped to SSN 1,2,...,9,0
  - Most of week 1 through week 3
- Off-Cycle
  - Days 22 through 2 of the next month
  - Week 4

---

### Emergency Assistance

Durham Social Services provides financial assistance for
- General emergency 
- Rent and utilities
- Food and transportation

### Cash Constrained

- Families coping with povery lack cash
- Need cash to buy food once SNAP runs out
- But cash is needed for other expenses
- **Demand for cash highest during off-cycle?**

---

# Data

### Durham Social Services (DSS)

- Emergency Assistance Requested & Provided
  - Date of request
  - Type of assistance
  - Case number and Names
- SNAP EBT
  - Transfer Dates
  - Amonts
  - Case number and Names

---

### Model

- Discrete-Time Events
- Benefits distribution date ($t=0$)

### Confounders

- Pay-period
- End-of-month cash demands
  - utilities
  - rent

---
class: middle, center

# Thank you